---
title: "Homework 1"
author: "Shay Culpepper"
date: "9/5/2018"
output: 
  html_document
  
references:
- id: fenner2012a
  title: One-click science marketing
  author:
  - family: Fenner
    given: Martin
  container-title: Nature Materials
  volume: 11
  URL: 'http://dx.doi.org/10.1038/nmat3283'
  DOI: 10.1038/nmat3283
  issue: 4
  publisher: Nature Publishing Group
  page: 261-263
  type: article-journal
  issued:
    year: 2012
    month: 3
---

---

### Study Group
* Kate
* Lara
* Lyzzy
* Flaka

### Pums Data
```{r setup, include=FALSE}
  setwd('/Users/shaypepper/Documents/school/econometrics')
  rm(list = ls(all = TRUE)) # clear workspace

  load("data/pums/pums_NY.RData")
  attach(dat_pums_NY)
  
  library(plyr)
```

#### Education by Borough

I need to work on tables. 
```{r}
  # dat_pums_NY$Borough <- 0
  # i = 0
  # 
  # for (b in c(in_Bronx, 
  #            in_Manhattan, 
  #            in_StatenI, 
  #            in_Brooklyn, 
  #            in_Queens, 
  #            in_Westchester)){
  #     dat_pums_NY$Borough <- dat_pums_NY$Borough + dat_pums_NY[b]*i
  #     i <- i + 1
  # }
  # table(Borough, row.names=c("Borough", "Education"))
    
```


### SP 500
This section is why my assignment is late. The good news that it has made me significantly more comfortable with R.

I went through a few iterations of this before I landed on an efficient algorithm that could handle any file from the [historical S&P data](https://www.investing.com/indices/us-spx-500-historical-data). I decided to look at ten years worth of data (09/05/2008-09/05/2018). That means the crash is included.

```{r}
setwd('/Users/shaypepper/Documents/school/econometrics')
rm(list = ls(all = TRUE)) # clear workspace
SPData <- read.csv(
  file="data/S&P500-tenyears.csv", 
  header=TRUE, 
  sep=",", 
  dec=".", 
  stringsAsFactors=FALSE)

names(SPData)[7] <- "Change"
N <- nrow(SPData)


# Order by date
SPData["Date"] <- as.Date(SPData$Date,
                          format("%b %d, %Y"))
attach(SPData)
SPData <- SPData[order(Date),]

# Convert Change string to numeric
SPData["Change"] <- as.numeric(gsub("%", "", SPData$Change))
```

My random variable, $X$, is consecutive days where $c = 1$.

$$ c_{n} = 
\begin{cases} 
  0,& \text{if } p_{n} \leq p_{n-1} \\
  1,& \text{otherwise}
\end{cases}
$$
$$ X = 
\begin{cases} 
  0,& \text{if } c_{n} =  0 \\
  |\{c_{i}: c_{i} = c_{i+1} = \ldots = c_{n} \}|,& \text{otherwise}
\end{cases}
$$

```{r}
positive <- SPData$Change > 0
streaks <- c(1:nrow(SPData))

for (i in streaks){
  tally <- 0
  for (j in i:1) {
    if (!positive[j]){
      break
    }
    tally <- tally + 1
  }
  streaks[i] <- tally
}
SPData$PositiveChange <- positive
SPData$NConsecutivePositives <- streaks

CalcData <- data.frame()
cumulative.total <- 0
m <- max(SPData$NConsecutivePositives)

# Loop backwards through all values of X
for (i in m:0){
  # Frequency of $X = i$
  n <- count(SPData$NConsecutivePositives == i)
  n <- n$freq[2]
  
  # Add n so smaller i's will be able to calculate proportions properly 
  cumulative.total <- cumulative.total + n
  
  row.data <- c(i, n, cumulative.total, 1 - n/cumulative.total)
  CalcData <- rbind(row.data, CalcData)
}

names(CalcData) <- c("X", "n", "cs", "p")
```

```{r echo=FALSE}
library(knitr)
kable(CalcData)
```

I honestly expected that this would show similar probabilities as $n$ got larger. It's probably just that cat study by [UK Observer](https://www.theguardian.com/money/2013/jan/13/investments-stock-picking). I got it in my head that stocks were mostly random. Instead, as n gets larger, the likelihood of another day of growth gets lower. I have a lot of questions about this, actually. I'll probably do some more diving into this at some point. Is there a similar effect for months? Hours? Does it work the opposite way as well after a series of losses? 

I probably could find a good paper on market corrections if I hadn't focused so much on the code. Next time!

Note: I don't remember whether or not you accept late homework, but even if you don't, I'd be interested in any comments on my code or method of tabulation that you may have. 






