---
title: "Exam 1 Practice Solutions"
subtitle: "Econ B2000, MA Econometrics"
author: "Kevin R Foster, CCNY"
date: "Fall 2018"
fig_width: 4
fig_height: 2 
output:
  html_document: default
  pdf_document: 
    toc: true
---
```{r echo = FALSE}
library(ggplot2)
density_plot <- function(lb, ub, mean = 0, sd = 1, dist.type = "normal", df = FALSE, limits = c(mean - 3 * sd, mean + 3 * sd)) {
    
  
    x <- seq(limits[1], limits[2], length.out = 100)
    xmin <- max(lb, limits[1])
    xmax <- min(ub, limits[2])
    areax <- seq(xmin, xmax, length.out = 100)
    
    
    if (dist.type == "normal") {
      y <- dnorm(x, mean = mean, sd = sd)
      ymax <- dnorm(areax, mean = mean, sd = sd)
    } else {
      y <- dt(x / sd, df = df)
      ymax <- dt(areax / sd, df = df)
    }
    
    area <- data.frame(x = areax, ymin = 0, ymax = ymax)
    
    (ggplot()
     + geom_line(data.frame(x = x, y = y),
                 mapping = aes(x = x, y = y))
     + geom_ribbon(data = area, mapping = aes(x = x, ymin = ymin, ymax = ymax))
     + scale_x_continuous(limits = limits)
     + theme(axis.title.x=element_blank(),
             axis.title.y=element_blank()))
}
```

```{r echo = FALSE, fig.height = 2, fig.width = 4}
density_plot(-Inf, 2, dist.type = "t", df = 1)
```

## Using the Normal and Student's T to find p-values
What will be helpful in this section: ``pnorm``, ``qnorm``, ``pt``, ``qt``, and a normal distribution / students t distribution graph to visualize. 2 ways to do it. 1) You can calculate the z score. $z = \frac{\bar{x} - \mu}{\sigma}$, or specify mean and sd in the function itself. pnorm and pt default to the lower tail. 


## Statistics from given numbers (no datasets in R required)

#### Population Mean
These assume an unknown sigma
$$
\begin{align}
  E = (t_{\alpha / 2}) \frac{s}{\sqrt{n}} 
  &&
  df = n - 1
  &&
  t = \frac{\bar{x} - \mu}{\frac{s}{\sqrt{n}}}
\end{align}
$$

#### Population proportion
$$
\begin{align}
  E &= z_{\alpha / 2} 
       \sqrt{\frac{\hat{p}(1 - \hat{p}) }{n} }
  &&
  z = \frac{\hat{p} - p}{\sqrt{ \frac{p(1-p)}{n} }}
  \\
  n &=  p(1-p)\left( \frac{z_{\alpha / 2}}{E}  \right)^2
\end{align}
$$


#### Difference in population means
These assume variances are different and unknown
$$
\begin{align}
  E &= 
    t_{\alpha / 2}
    \sqrt{
      \frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}
    }
  &&
  df = min(n_1, n_2) - 1 
  \\
  \\
  t &= \frac{
      \left(\bar{x}_1 - \bar{x}_2 \right) - 
      \left(\mu_1 - \mu_2 \right)}
    {\sqrt{
      \frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}
    }}
\end{align}
$$

#### Difference in population proportions

$$
\begin{align}
  E &= z_{\alpha / 2} \sqrt { 
    \frac{ \hat{p}_1 (1 - \hat{p}_1) }{n_1} + \frac{ \hat{p}_2 (1 - \hat{p}_2)}{n_2} 
    }
  &&
  \bar{p} = \frac{x_1 + x_2}{n_1 + n_2}
  \\
  \\
  z &= \frac{
  \left(\hat{p}_1 - \hat{p}_2 \right) - 
  \left(p_1 - p_2 \right)}
  {\sqrt{
    \bar{p} \left( 1 - \bar{p}\right) \left( \frac{1}{n_1} + \frac{1}{n_2} \right)
  }}
\end{align}
$$

## Regression Analysis from given data (no datasets in R required)

## Statistics using Datasets (R required)

## Regression Analysis using Datasets (R Required)
