---
title: "Exam 1 Cheatsheet"
subtitle: "Econ B2000, MA Econometrics"
author: "Shay Culpepper, CCNY"
date: "Fall 2018"
fig_width: 4
fig_height: 2 
output:
  pdf_document: default
  html_document: default
---
```{r echo = FALSE}
library(ggplot2)
density_plot <- function(lb, ub, mean = 0, sd = 1, dist.type = "normal", df = FALSE, limits = c(mean - 3 * sd, mean + 3 * sd)) {
    
  
    x <- seq(limits[1], limits[2], length.out = 100)
    xmin <- max(lb, limits[1])
    xmax <- min(ub, limits[2])
    areax <- seq(xmin, xmax, length.out = 100)
    
    
    if (dist.type == "normal") {
      y <- dnorm(x, mean = mean, sd = sd)
      ymax <- dnorm(areax, mean = mean, sd = sd)
    } else {
      y <- dt(x / sd, df = df)
      ymax <- dt(areax / sd, df = df)
    }
    
    area <- data.frame(x = areax, ymin = 0, ymax = ymax)
    
    (ggplot()
     + geom_line(data.frame(x = x, y = y),
                 mapping = aes(x = x, y = y))
     + geom_ribbon(data = area, mapping = aes(x = x, ymin = ymin, ymax = ymax))
     + scale_x_continuous(limits = limits, breaks = c(lb, ub, mean))
     + theme(axis.title.x=element_blank(),
             axis.title.y=element_blank(),
             axis.ticks.y=element_blank(),
             axis.text.y = element_blank()
             ))
}
```



## Using the Normal and Student's T to find p-values
What will be helpful in this section: ``pnorm``, ``qnorm``, ``pt``, ``qt``, and a normal distribution / students t distribution graph to visualize. 2 ways to do it. 1) You can calculate the z score. $z = \frac{\bar{x} - \mu}{\sigma}$, or specify mean and sd in the function itself. pnorm and pt default to the lower tail. 


``pt(2, df = 10)`` gives you the area of the shaded region in the figure below.

```{r echo = FALSE, fig.height = 0.75, fig.width = 3}
density_plot(-Inf, 2, dist.type = "t", df = 10)
```

``pt(2, df = 10, lower.tail = FALSE)`` gives you the area of the shaded region in the figure below.

```{r echo = FALSE, fig.height = 0.75, fig.width = 3}
density_plot(2, Inf, dist.type = "t", df = 10)
```

``pnorm(0.5)`` gives you the area of the shaded region in the figure below.

```{r echo = FALSE, fig.height = 0.75, fig.width = 3}
density_plot(-Inf, 0.5, dist.type = "normal")
```

``pnorm(0.5, mean = 0.6, sd = 0.1)`` gives you the area of the shaded region in the figure below.

```{r echo = FALSE, fig.height = 0.75, fig.width = 3}
density_plot(-Inf, 0.5, mean = 0.6, sd = 0.1, dist.type = "normal")
```

``pnorm(1.67, lower.tail = FALSE)`` gives you the area of the shaded region in the figure below.

```{r echo = FALSE, fig.height = 0.75, fig.width = 3}
density_plot(1.67, Inf, dist.type = "normal")
```


``pnorm(2) - pnorm(-2)`` gives you the area of the shaded region in the figure below.

```{r echo = FALSE, fig.height = 0.75, fig.width = 3}
density_plot(-2, 2, dist.type = "normal")
```

\newpage

## Statistics from given numbers (no datasets in R required)

#### Means
These assume an unknown sigma
\begin{align} \nonumber
  E = (t_{\alpha / 2}) \frac{s}{\sqrt{n}} &&
  df = n - 1 &&
  t = \frac{\bar{x} - \mu}{\frac{s}{\sqrt{n}}} &&
  n = \left( \frac{ z_{\alpha / 2} s }{E} \right)^2 &&
  SE = \frac{s}{\sqrt{n}}
\end{align}


##### Difference in means
These assume variances are different and unknown

\begin{align} \nonumber
  SE = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}} && 
  E = 
    t_{\alpha / 2} SE &&
  df = min(n_1, n_2) - 1  &&
  t = \frac{
      \left(\bar{x}_1 - \bar{x}_2 \right) - 
      \left(\mu_1 - \mu_2 \right)}
    {\sqrt{
      \frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}
    }}
\end{align}

#### Population proportion

\begin{align} \nonumber
 SE = \sqrt{\frac{\hat{p}(1 - \hat{p}) }{n} } &&
  E = z_{\alpha / 2} SE &&
  z = \frac{\hat{p} - p}{\sqrt{ \frac{p(1-p)}{n} }} &&
  n =  p(1-p)\left( \frac{z_{\alpha / 2}}{E}  \right)^2
\end{align}

#### Difference in proportions

\begin{align} \nonumber
  SE = \sqrt { 
    \frac{ \hat{p}_1 (1 - \hat{p}_1) }{n_1} + \frac{ \hat{p}_2 (1 - \hat{p}_2)}{n_2} 
    } &&
  E = z_{\alpha / 2} SE &&
  \bar{p} = \frac{x_1 + x_2}{n_1 + n_2} &&
  z = \frac{
  \left(\hat{p}_1 - \hat{p}_2 \right) - 
  \left(p_1 - p_2 \right)}
  {\sqrt{
    \bar{p} \left( 1 - \bar{p}\right) \left( \frac{1}{n_1} + \frac{1}{n_2} \right)
  }} 
\end{align}

#### Hypothesis Test

\begin{align} 
\nonumber H_0         && H_a         && Test        && Reject\ H_0 if...          \\
\nonumber \mu \geq a  && \mu < a     && left-tail   && test\ stat < -Z_{\alpha}   && or\ p < \alpha \\
\nonumber \mu \leq a  && \mu > a     && right-tail  && test\ stat >  Z_{\alpha}   && or\ p < \alpha \\
\nonumber \mu = a     && \mu \neq a  && two-tail    && |test\ stat| > Z_{\alpha / 2} && or\ p < \alpha
\end{align}

## Regression Analysis from given data (no datasets in R required)

\begin{align}
  test\ stat = \frac{\beta_i}{SE_i} &&
  SE_i = \frac{\beta_i}{test\ stat} &&
  \beta_i = (test\ stat) SE
\end{align}

To caluclate p-value, use student's t if you have n or are given the degrees of freedom explicitly. Generally, your degrees of freedom for a beta coefficient should be $n - p$ with n being your sample size, and p being the number of parameters being predicted including the constant coefficient. If you haven't been given degrees of freedom or the sample size, you are probably safe to assume that using the normal would be sufficient. 

#### Regularize data
```{r}
M <- max(x,na.rm = TRUE)
m <- min(x,na.rm = TRUE)

norm_varb <- function(x) { (x-m)/abs(M-m) }
```

## Statistics using Datasets (R required)

## Regression Analysis using Datasets (R Required)
