---
title: "Regression types"
author: "Shay Culpepper"
date: "11/13/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

............

```{r eval=FALSE}

library('rpart') # tree
library('randomForest') # random forest

```

## 1. Basic OLS
Our most basic regression. 
### Formula
### Code in R
### Example interpretation

## 2. Logit
### Formula
### Code in R
### Example interpretation

## 3. Probit
### Formula
### Code in R
### Example interpretation

## 4. k-nn
### Formula
### Code in R
### Example interpretation

## 5. Interactions/multi-level
### Formula
### Code in R
### Example interpretation

## 6. Quantile Regression
Sometimes it is less relevant to look at all off your data in a single regression. There are times that you are interested in the effect of your exogenous variables based on being part of a specific quantile. 
### Formula
### Code in R
```{r eval=FALSE}
p_tiles <- c(0.1, 0.25, 0.5, 0.75, 0.9)
quantreg1 <- rq(WSAL_VAL ~ Age + female + AfAm + Asian + Amindian + race_oth + Hispanic + educ_hs + educ_smcoll + educ_as + educ_bach + educ_adv + married + divwidsep + union_m + veteran + immigrant + immig2gen, tau=p_tiles, data = dat_use)
summary(quantreg1)
plot(quantreg1)
```
### Example interpretation

## 7. LOESS
Local estimation with polynomials. Maximum 4 variables. 
### Formula
### Code in R

```{r eval=FALSE}
model_loess1 <- loess(WSAL_VAL ~ Age,data3)
y_loess1_pred <- predict(model_loess1, data.frame(Age = seq(25, 55, 1)), se = TRUE)
plot(seq(25, 55, 1),y_loess1_pred$fit)
```

### Example interpretation

## 8. Propensity Score
### Formula
### Code in R
### Example interpretation

## 9. LASSO
### Formula
### Code in R
### Example interpretation

## 10. Spike and Slab
### Formula
### Code in R
### Example interpretation

## 11. Random Forrest
### Formula
### Code in R
### Example interpretation

## 12. Tree
### Formula
### Code in R
```{r eval = FALSE}

```
### Example interpretation


## 13. Non-parametric
### Formula
### Code in R
### Example interpretation

## 14. Spline
### Formula
### Code in R
### Example interpretation


