---
title: "Homework 2"
author: "Ekaterina Marbot, Elizabeth Chung, Shay Culpepper, Flaka Bajraktari"
date: "9/25/2018"
output:
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(plyr)

setwd("/Users/shaypepper/Documents/school/econometrics")
load("data/pums/pums_NY.RData")



```

## Benford's Law

```{r}
library(benford.analysis)
data(corporate.payment)
bfd.cp <- benford(corporate.payment$Amount)
plot(bfd.cp)
```

``` {r}
bfd.cp
```

## Education and Income
Proportion of people with less than a college degree earn less than $70,000 income for each borough

#### Bronx
```{r}
prop.table(table(dat_pums_NY$in_Bronx, dat_pums_NY$educ_nohs, dat_pums_NY$income_total <70000))
```

It means that proportion of people without high school degree earns less than $70,000 income in Bronx is 0.0264.
 
We can find the same result by using subset and display it for each borough.

```{r}
dat_use1 <- subset(dat_pums_NY,((income_total < 70000) & educ_nohs))
```

Then use function ddply.

Proportion of people without high school degree earns less than $70,000 income in Staten Island - 0.0062807543, in in Brooklyn - 0.0476023106, in Queens - 0.037399269. 
 
Proportion of people without high school degree earns more than $70,000 income in Bronx, for example - 0.0001426286.
 
Regarding proportion of people with college degree, for example, how have 
Bachelor degree, in Manhattan is 0.006912395 each less than $70,000, and 0.004706745 how 
earn more than $70,000. 


## Income inequality, averages, by neighborhood and borough
This isn't one of your questions, but this boxplot is a nice representation of the inequality by borough.
```{r}

pums.compact <- dat_pums_NY %>%
  mutate(borough = factor(
           1 * in_Bronx + 
           2 * in_Brooklyn + 
           3 * in_Manhattan + 
           4 * in_StatenI + 
           5 * in_Queens + 
           6 * in_Westchester + 
           7 * in_Nassau, 
           levels = 0:7,
           labels = c("Not NYC", 
                      "Bronx", 
                      "Brooklyn", 
                      "Manhattan", 
                      "Staten Island", 
                      "Queens", 
                      "Westchester", 
                      "Nassau")
           ),
         edu = ordered(
           1 * educ_nohs + 
             2 * educ_hs + 
             3 * educ_smcoll + 
             4 * educ_as + 
             5 * educ_bach +  
             6 * educ_adv,
           levels = 1:6,
           labels = c("No High School", 
                      "High School", 
                      "Some College", 
                      "Associates", 
                      "Bachelors", 
                      "Advanced Degree")
           ),
        race = factor(
          1 * white + 
          1 * Hispanic + 
          3 * AfAm + 
          4 * Amindian + 
          5 * Asian + 
          6 * race_oth,
          levels = 1:6,
          labels = c("White - non-hispanic", 
                     "White - hispanic",
                     "African American",
                     "American Indian",
                     "Asian",
                     "Other")
          ), 
        PUMA = factor(PUMA, 
                      levels=c(03701, 03702, 03703, 03704, 03705, 03706, 
                               03707, 03708, 03709, 03710, 03801,
                               03802, 03803, 03804, 03805, 03806, 
                               03807, 03808, 03809, 03810, 03901,
                               03902, 03903, 04001, 04002, 04003, 
                               04004, 04005, 04006, 04007, 04008,
                               04009, 04010, 04011, 04012, 04013,
                               04014, 04015, 04016, 04017, 04018, 
                               04101, 04102, 04103, 04104, 04105, 
                               04106, 04107, 04108, 04109, 04110,
                               04111, 04112, 04113, 04114),
                      labels= c("Riverdale, Fieldston & Kingsbridge",
                                "Wakefield, Williamsbridge & Woodlawn",
                                "Co-op City, Pelham Bay & Schuylerville",
                                "Pelham Parkway, Morris Park & Laconia",
                                "Belmont, Crotona Park East & East Tremont",
                                "Bedford Park, Fordham North & Norwood",
                                "Morris Heights, Fordham South & Mount Hope",
                                "Concourse, Highbridge & Mount Eden",
                                "Castle Hill, Clason Point & Parkchester",
                                "Hunts Point, Longwood & Melrose",
                                "Washington Heights, Inwood & Marble Hill",
                                "Hamilton Heights, Manhattanville & West Harlem",
                                "Central Harlem",
                                "East Harlem",
                                "Upper East Side",
                                "Upper West Side & West Side",
                                "Chelsea, Clinton & Midtown Business District",
                                "Murray Hill, Gramercy & Stuyvesant Town",
                                "Chinatown & Lower East Side",
                                "Battery Park City, Greenwich Village & Soho",
                                "Tottenville, Great Kills & Annadale",
                                "New Springville & South Beach",
                                "Port Richmond, Stapleton & Mariner's Harbor",
                                "Greenpoint & Williamsburg",
                                "Bushwick",
                                "Bedford-Stuyvesant",
                                "Brooklyn Heights & Fort Greene",
                                "Park Slope, Carroll Gardens & Red Hook",
                                "Crown Heights North & Prospect Heights",
                                "Brownsville & Ocean Hill",
                                "East New York & Starrett City",
                                "Canarsie & Flatlands",
                                "East Flatbush, Farragut & Rugby",
                                "Crown Heights South, Prospect Lefferts & Wingate",
                                "Sunset Park & Windsor Terrace",
                                "Bay Ridge & Dyker Heights",
                                "Borough Park, Kensington & Ocean Parkway",
                                "Flatbush & Midwood",
                                "Sheepshead Bay, Gerritsen Beach & Homecrest",
                                "Bensonhurst & Bath Beach",
                                "Brighton Beach & Coney Island",
                                "Astoria & Long Island City",
                                "Jackson Heights & North Corona",
                                "Flushing, Murray Hill & Whitestone",
                                "Bayside, Douglaston & Little Neck",
                                "Queens Village, Cambria Heights & Rosedale",
                                "Briarwood, Fresh Meadows & Hillcrest",
                                "Elmhurst & South Corona",
                                "Forest Hills & Rego Park",
                                "Sunnyside & Woodside",
                                "Ridgewood, Glendale & Middle Village",
                                "Richmond Hill & Woodhaven",
                                "Jamaica, Hollis & St. Albans",
                                "Howard Beach & Ozone Park",
                                "Far Rockaway, Breezy Point & Broad Channel)"
                      ))
      ) %>%
  select(`borough`, `edu`, `race`, PUMA, income_total, Age) %>%
  filter(borough != "Not NYC", income_total >= 0, Age >= 18) %>%
  arrange(PUMA)



pums.income_by_neighborhood <- pums.compact %>%
  filter(borough != "Westchester" & borough != "Nassau") %>%
  group_by(`borough`, PUMA) %>%
  summarize(meanIncome = mean(income_total), 
            q75.less.q25 = quantile(income_total, 0.75) - quantile(income_total, 0.25),
            q10 = quantile(income_total, 0.10), 
            q25 = quantile(income_total, 0.25), 
            q75 = quantile(income_total, 0.75), 
            q90 = quantile(income_total, 0.90),
            sd = sd(income_total)
            )


ggplot(pums.income_by_neighborhood, aes(x = borough, y = "meanIncome")) +
  geom_boxplot()

```

### Richest and poorest neighborhoods by borough

#### Highest Mean Income
```{r, echo=FALSE}
pums.highest.income.by.borough <- pums.income_by_neighborhood %>%
  top_n(n=1, meanIncome) %>%
  select(borough, PUMA, meanIncome)

pums.highest.income.by.borough
```
#### Lowest Mean Income
```{r, echo=FALSE}
pums.lowest_income_by_borough <- pums.income_by_neighborhood %>%
  top_n(n=1, -meanIncome) %>%
  select(borough, PUMA, meanIncome)
  
pums.lowest_income_by_borough
```

### Inequality by neighborhood

#### Measured by 75 quartile minus 25 quartile.

```{r, echo=FALSE}
pums.most.unequal.by.iqr <- pums.income_by_neighborhood %>%
  top_n(n=1, q75.less.q25) %>%
  select(borough, PUMA, q75.less.q25)

pums.most.unequal.by.iqr
```
______

#### Measured by standard deviation
```{r, echo=FALSE}

pums.most.unequal.by.sd <- pums.income_by_neighborhood %>%
  top_n(n=1, sd) %>%
  select(borough, PUMA, sd)

pums.most.unequal.by.sd

```
_______

#### Compare differences between approaches

```{r, echo=FALSE}
top.by.sd <- pums.income_by_neighborhood %>%
  filter(borough == "Manhattan") %>%
  arrange(-sd) %>%
  select(borough, PUMA, sd)

top.by.iqr <- pums.income_by_neighborhood %>%
  filter(borough == "Manhattan") %>%
  arrange(-q75.less.q25) %>%
  select(borough, PUMA, q75.less.q25)

head(top.by.iqr)

head(top.by.sd)
```

## Taxi Data
Subsample of the taxi data:
rides paid with credit cards - 193750 rides 
rides paid with cash – 187694 rides

The point estimate:
Sample proportion = (rides paid in cash)/(sample size).
Total number of ride $n = 381444$
Sample proportion = 0.4921

The margin of error: 

$$
c = 0.9 \\
α = 1 - c = 0.1 \\
z_{\frac{\alpha}{2}} =  z_{0.05} = 1.645 \\
E = 0.0013
$$

Lower and Upper endpoint:
Lower endpoint $= p\hat = 0.4907$
Upper endpoint = 0.4934
90% confidence interval for the fraction of rides paid in cash on the weekend: 
$0.4907 < p < 0.4934$



